A full approximation scheme multilevel method \\ for solving nonlinear variational inequalities

### 2024-07-21 talk-ismp/slides.tex
work some on the talk

\titlegraphic{\vspace{-8mm} \includegraphics[width=0.26\textwidth]{../talk-oxford/images/uafbw.png} \hfill \begin{minipage}{0.14\textwidth}
\includegraphics[width=\textwidth]{../talk-oxford/images/oxford.png}

@@

{
  %\usebackgroundtemplate{\includegraphics[width=\paperwidth]{../talk-oxford/images/gray-british-clark2022.png}}
  \begin{frame}
    \titlepage
  \end{frame}
}

\begin{frame}{Outline}
  \tableofcontents[hideallsubsections]
\end{frame}

@@

\begin{center}
\begin{tabular}{r|l|l}
& \qquad unconstrained & \qquad constrained \\ \hline
optimization &
\begin{minipage}[t][16mm][t]{0.32\textwidth}
$$\min_{u\in\mathcal{V}} J(u)$$
\end{minipage}
&
\begin{minipage}[t][16mm][t]{0.35\textwidth}
$$\min_{u\in\mathcal{K}} J(u)$$
\end{minipage}
\\ \hline
\only<1>{equations}\only<2>{\begin{minipage}[t][16mm][t]{0.15\textwidth} weak form \par equations \end{minipage}} &
\begin{minipage}[t][16mm][t]{0.32\textwidth}

\vspace{-2mm}
find $u \in \mathcal{V}$:
\only<1>{$$F(u)=0$$}
\only<2>{$$\ip{F(u)}{v} = 0 \quad \forall v \in \mathcal{V}$$}
\end{minipage}
&
\begin{minipage}[t][16mm][t]{0.35\textwidth}

\vspace{-2mm}
\only<2>{
{\color{FireBrick} find $u \in \mathcal{K}$:}
$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \forall v \in \mathcal{K}}$$
}
\end{minipage}
\end{tabular}
\end{center}

@@

\begin{itemize}
\item where $\mathcal{K} \subset \mathcal{V}$ is closed and convex
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \emph{problem.} on $\Omega \subset \RR^2$, find $u(x)$ which minimizes
    $$J(v) = \int_\Omega \frac{1}{2} |\grad v|^2 - f\, v$$
over $\mathcal{K} = \left\{v \in H^1(\Omega) \,:\, v\big|_{\partial \Omega} = g \,\text{ and }\, v \ge \psi\right\}$
\end{itemize}
\end{frame}

\includegraphics[width=0.55\textwidth]{../talk-oxford/images/obstacle65.pdf} \qquad \includegraphics[width=0.35\textwidth]{../talk-oxford/images/obstacle-sets.png}

\bigskip
\begin{itemize}
\item the solution defines subsets of $\Omega$:
   \begin{itemize}
   \item[$\circ$] \emph{active set} $A_u = \{u = \psi\}$
   \item[$\circ$] \emph{inactive set} $R_u = \{u> \psi\}$
   \item[$\circ$] \emph{free boundary} $\Gamma_u=\partial R_u \cap \Omega$
   \end{itemize}
\item the \emph{complementarity problem} (CP) is a PDE-like strong form:
\begin{equation*}
u - \psi \ge 0, \qquad -\grad^2 u - f \ge 0, \qquad (u - \psi)(-\grad^2 u - f) = 0
\end{equation*}

   \begin{itemize}
   \item[$\circ$] CP $=$ KKT conditions essentially, but $\infty$-dimensional
   \end{itemize}
\item the weak form is a {\color{FireBrick} \emph{variational inequality} (VI)}, which says that $J'(u)$ points directly into $\mathcal{K}$:
    $$\ip{J'(u)}{v-u} = \int_\Omega \grad u\cdot \grad (v-u) - f (v-u) \ge 0$$
for all $v \in \mathcal{K}$
\end{itemize}
\end{frame}

\begin{frame}{VI $=$ weak form}

\begin{itemize}
\item for problems of optimization type, the VI is the weak form, with $v-u$ as the test function:
{\small
   $$J(u) \le J(v) \quad \forall v \in \mathcal{K} \qquad \iff \qquad \ip{J'(u)}{v-u} \ge 0 \quad \forall v \in \mathcal{K}$$
}
\end{itemize}

\bigskip
\begin{center}
\includegraphics[width=0.5\textwidth]{../talk-dms/figs/convexuv.png}
\end{center}
\end{frame}

@@

\begin{itemize}
\item let $\mathcal{K}$ be a closed and convex subset of a Banach space $\mathcal{V}$
\item suppose $F:\mathcal{K} \to \mathcal{V}'$ is a continuous operator
    \begin{itemize}
    \item[$\circ$] $F$ is generally nonlinear
    \item[$\circ$] $F$ may be defined \emph{only} on $\mathcal{K}$
    \item[$\circ$] $F$ may \emph{not}\, be the derivative of an objective function $J$
    \item[$\circ$] $F=J'$, a linear operator, in classical obstacle problem
    \end{itemize}
\item the general variational inequality {\color{FireBrick} VI($F$,$\mathcal{K}$)} is
	$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \text{ for all } v \in \mathcal{K}}$$
\item when $\mathcal{K}$ is nontrivial the problem {\color{FireBrick} VI($F$,$\mathcal{K}$)} is nonlinear \emph{even when $F$ is a linear operator}
\end{itemize}
\end{frame}

@@

\vspace{3mm}
\item first-semester calculus:
    $$u \gets \min_{x\in[a,b]} f(x) \quad \iff \quad f'(u)(v-u) \ge 0 \quad \forall v \in[a,b]$$
\end{itemize}

%\vspace{20mm}
\begin{center}
\includegraphics[width=0.9\textwidth]{../talk-oxford/images/calcone.png}
\end{center}
\end{frame}

@@

\section{nonlinear (FAS) multigrid for PDEs}

\begin{frame}{nonlinear 2-mesh scheme}

@@

	\begin{itemize}
	\item[$\circ$] $u\in \mathcal{V}=H^1(\Omega)$
	\item[$\circ$] $\ell\in \mathcal{V}'$
	\item[$\circ$] $F : \mathcal{V} \to \mathcal{V}'$ continuous and one-to-one
	\item[$\circ$] for example, the Liouville-Bratu problem: $-\grad^2 u - e^u = f$
	\end{itemize}
\item discretization gives algebraic system on fine mesh $\Omega^h$:
    $$F^h(u^h) = \ell^h$$

	\begin{itemize}
	\item[$\circ$] $u^h$ denotes exact (algebraic) solution
	\end{itemize}
\end{itemize}

@@

	\begin{itemize}
	\item are there alternatives to Newton's method?
	\end{itemize}
\item notes:
    \setbeamertemplate{enumerate item}{\emph{\roman{enumi})}}
    \begin{enumerate}
    \item the \emph{residual} \quad $r^h = \ell^h - F^h(w^h)$ \quad is computable
    \item the \emph{error} \quad $e^h = u^h-w^h$ \quad is unknown
    \item our equation can be rewritten
    $$F^h(u^h) - F^h(w^h) = r^h$$
    \end{enumerate}
\end{itemize}
}
\only<3-4>{
\begin{itemize}
\item \emph{updated goal}: from iterate $w^h$, to solve
    $$F^h(u^h) - F^h(w^h) = r^h$$
\item \alert{for $F^h$ linear}, convert this to the \emph{error equation}
    $$F^h(e^h) = r^h$$
\item an approximation solution $\tilde e^h$ would improve our iterate:
    $$w^h \leftarrow w^h+\tilde e^h$$
\item<4> but $F^h$ is not linear!
\end{itemize}

\vspace{5mm}
}
\only<5>{
\begin{itemize}
\item \emph{updated goal}: use a coarser mesh $\Omega^H$ to somehow estimate the solution $u^h$ in the nonlinear \emph{correction equation}
    $${\color{FireBrick} F^h(u^h) - F^h(w^h) = r^h}$$
\item basic multigrid idea: there are algorithms (\alert{smoothers}) which ``improve'' $w^h$ \dots use them a little first \dots then correct from the coarser mesh
    \begin{itemize}
    \item[$\circ$] ``improve'' means they remove high-frequency error components efficiently
    \end{itemize}
\end{itemize}

\vspace{15mm}
}
\only<6>{
\begin{itemize}
\item \emph{nodewise problem}: for $\psi_i^h$ a hat function or dof, solve for $c\in\RR$ to make the residual at that location zero:
	$${\color{FireBrick} \phi_i(c) = r^h(w^h + c \psi_i^h)[\psi_i^h] = 0}$$
\item sweeping through and solving nodewise problems is a \alert{smoother}
    \begin{itemize}
    \item[$\circ$] Fourier analysis shows smoothing property
    \item[$\circ$] after smoothing, $e^h$ and $r^h$ have smaller high-frequencies
    \end{itemize}
\item after smoothing, the correction equation on $\Omega^h$ should be accurately approximate-able on the coarser mesh $\Omega^H$
\end{itemize}

\vspace{14mm}
}
\only<7>{
\begin{itemize}
\item \emph{updated goal}: use a coarser mesh $\Omega^H$ to somehow estimate the solution $u^h$ in $F^h(u^h) - F^h(w^h) = r^h(w^h)$
\item Brandt's (1977) \emph{full approximation scheme} (FAS) equation:
	$${\color{FireBrick} F^H(u^H) - F^H(\iR w^h) = R \, r^h(w^h)}$$

@@

\item rewritten FAS equation: let ${\color{FireBrick} \ell^H = F^H(\iR w^h) + R\, r^h(w^h)}$ then
    $${\color{FireBrick} F^H(u^H) = \ell^H}$$
\end{itemize}
}
\end{frame}

\begin{frame}{full approximation scheme (FAS) 2-mesh solver}

@@

\begin{align*}
&\text{pre-smooth over fine:} & & [\text{smoother updates } w^h] \\
&\text{restrict:}                   & &\ell^H = F^H(\iR w^h) + R\, r^h(w^h) \\
&\text{solve coarse:}                      & &F^H(w^H) = \ell^H \\
&\text{correct:}                    & &w^h \leftarrow w^h + P(w^H - \iR w^h) \\
&\text{post-smooth over fine:} & & [\text{smoother updates } w^h]
\end{align*}

\bigskip
{\small
\begin{itemize}
\item $P: \mathcal{V}^H \to \mathcal{V}^h$ is \emph{canonical prolongation}
\item \textbf{restrict}$+$(\textbf{solve coarse})$+$\textbf{correct} \, $=$ \, \emph{FAS coarse grid correction}
\end{itemize}
}
\end{frame}

@@

\begin{frame}{does it work?}

\begin{itemize}
\item FAS multigrid works \alert{very well} on nice nonlinear PDE problems
\item example: Liouville-Bratu equation
    $$-\nabla^2 u - e^u = 0$$
with Dirichlet boundary conditions on $\Omega=(0,1)^2$
\item discretize by (straightforward) finite differences
\item minimal problem-specific code:
    \begin{enumerate}
    \item[1.] residual evaluation on grid level: $F^j(\cdot)$
    \item[2.] pointwise smoother: $\phi_i(c) = 0 \,\forall i$
        \begin{itemize}
        \item[$\circ$] nonlinear Gauss-Seidel iteration
        \end{itemize}
    \item[3.] coarsest-level solve can be same as smoother, or more sophisticated (e.g.~Newton iteration)
    \end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}{the meaning of ``fast solver''}

\begin{itemize}
\item what does ``very well'' on the previous slide mean?
\end{itemize}

\begin{block}{definition} a solver is \emph{optimal} if work in flops, and/or run-time, is $O(N)$ for $N$ unknowns
\end{block}

\begin{itemize}
\item since $\sim$1980: optimality can be achieved by multigrid for PDE problems with reasonably-smooth solutions
\item in fact, multigrid people get greedy
\begin{block}{definition} a solver shows \emph{textbook multigrid efficiency} if it does total work less than 10 times that of a single smoother sweep
\end{block}
    \begin{itemize}
    \item[$\circ$] TME $\implies$ optimal
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Bratu model problem: TME}

\begin{columns}
\begin{column}{0.45\textwidth}
\begin{itemize}
\item \texttt{bratu.c}
\item observed optimality:
\begin{align*}
\text{flops} &= O(N^1) \\
\text{processor time} &= O(N^1)
\end{align*}
\item highest-resolution {\color{FireBrick} $12$-level V-cycle} has $N\approx 10^8$ unknowns
\item compare $\approx 20\,\mu\,\text{s}/\text{N}$ for Poisson using Firedrake ($P_1$, geometric multigrid)
\end{itemize}
\end{column}
\begin{column}{0.55\textwidth}
\includegraphics[width=\textwidth]{../talk-dms/figs/bratu-time.png}
\end{column}
\end{columns}
\end{frame}

\section{FASCD = FAS multigrid for VIs}

@@

\section{results}

>>>>

\titlegraphic{\vspace{-18mm} \includegraphics[width=0.22\textwidth]{../talk-oxford/images/uafbw.png} \hfill \begin{minipage}{0.12\textwidth}
\includegraphics[width=\textwidth]{../talk-oxford/images/oxford.png}

@@

{
  %\usebackgroundtemplate{\includegraphics[width=\paperwidth]{../talk-oxford/images/gray-british-clark2022.png}}
  \begin{frame}

\vspace{10mm}

    \titlepage
  \end{frame}
}

\begin{frame}{Outline}
  \tableofcontents
\end{frame}

@@

\begin{center}
\begin{tabular}{r|l|l}
& \qquad unconstrained & \qquad constrained \\ \hline
optimization &
\begin{minipage}[t][16mm][t]{0.32\textwidth}
$$\min_{u\in\mathcal{V}} J(u)$$
\end{minipage}
&
\begin{minipage}[t][16mm][t]{0.35\textwidth}
$$\min_{u\in\mathcal{K}} J(u)$$
\end{minipage}
\\ \hline
\only<1>{equations}\only<2->{\begin{minipage}[t][16mm][t]{0.15\textwidth} weak form \par equations \end{minipage}} &
\begin{minipage}[t][16mm][t]{0.32\textwidth}

\vspace{-2mm}
find $u \in \mathcal{V}$:
\only<1>{$$F(u)=0$$}
\only<2->{$$\ip{F(u)}{v} = 0 \quad \forall v \in \mathcal{V}$$}
\end{minipage}
&
\begin{minipage}[t][16mm][t]{0.35\textwidth}

\vspace{-2mm}
\only<3>{
{\color{FireBrick} find $u \in \mathcal{K}$:}
$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \forall v \in \mathcal{K}}$$
}
\end{minipage}
\end{tabular}
\end{center}

@@

\begin{itemize}
\item where $\mathcal{K} \subset \mathcal{V}$ is closed and convex
\item<3> sometimes the {\color{FireBrick} variational inequality (VI)} is the KKT condition of optimization ($F=J'$), and sometimes not
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item \emph{problem.} on $\Omega \subset \RR^2$, find $u(x)$ which minimizes
    $$J(v) = \int_\Omega \frac{1}{2} |\grad v|^2 - f\, v$$
over
    $$\mathcal{K} = \left\{v \in H^1(\Omega) \,:\, v\big|_{\partial \Omega} = g \,\text{ and }\, v \ge \psi\right\}$$
\end{itemize}
\end{frame}

\begin{frame}{example, cont.}

\begin{minipage}[t]{0.55\textwidth}
\vspace{0pt}
\includegraphics[width=\textwidth]{../talk-oxford/images/obstacle65.pdf}
\end{minipage}
\hfill
\begin{minipage}[t]{0.3\textwidth}
\vspace{2mm}
\includegraphics[width=\textwidth]{../talk-oxford/images/obstacle-sets.png}
\end{minipage}

\vspace{-1mm}
\begin{itemize}
\item the solution defines subsets of $\Omega$:
   \begin{itemize}
   \item[$\circ$] \emph{active set} $A_u = \{u = \psi\}$
   \item[$\circ$] \emph{inactive set} $R_u = \{u> \psi\}$
   \item[$\circ$] \emph{free boundary} $\Gamma_u=\partial R_u \cap \Omega$
   \end{itemize}
\item the \emph{complementarity problem} (CP) is a PDE-like strong form:
\begin{equation*}
u - \psi \ge 0, \qquad -\grad^2 u - f \ge 0, \qquad (u - \psi)(-\grad^2 u - f) = 0
\end{equation*}
\item the weak form is the VI:
    $$\int_\Omega \grad u\cdot \grad (v-u) - f (v-u) \ge 0 \quad \forall v \in \mathcal{K}$$
\end{itemize}
\end{frame}

\begin{frame}{geometrical meaning of VI}

\begin{itemize}
\item the VI gives an ``angle condition'' for the test function $v-u$:
\item in optimization problems the condition is a consequence:
\begin{align*}
J(u) &\le J(v) \quad \forall v \in \mathcal{K} \\
\implies \qquad \ip{J'(u)}{v-u} &\ge 0 \quad \forall v \in \mathcal{K}
\end{align*}
\item for general VIs the condition \emph{is} the problem:
   $$\ip{F(u)}{v-u} \ge 0 \quad \forall v \in \mathcal{K}$$    
\end{itemize}

\bigskip
\begin{center}
\includegraphics[width=0.4\textwidth]{../talk-dms/figs/convexuv.png}
\end{center}
\end{frame}

@@

\begin{itemize}
\item let $\mathcal{K}$ be a closed and convex subset of a Banach space $\mathcal{V}$
\item suppose $F:\mathcal{K} \to \mathcal{V}'$ is a continuous operator
    \begin{itemize}
    \item[$\circ$] $F$ is generally nonlinear
    \item[$\circ$] $F$ may be defined \emph{only} on $\mathcal{K}$
    \item[$\circ$] $F$ may \emph{not}\, be the derivative of an objective function $J$
    \end{itemize}
\item the general variational inequality {\color{FireBrick} VI($F$,$\mathcal{K}$)} is
	$${\color{FireBrick} \ip{F(u)}{v-u} \ge 0 \quad \text{ for all } v \in \mathcal{K}}$$
\item when $\mathcal{K}$ is nontrivial, VI($F$,$\mathcal{K}$) is nonlinear even when $F$ is a linear operator
\end{itemize}
\end{frame}

@@

\vspace{1.5mm}
\item<2> first-semester calculus:
    $$u = \mathop{\textnormal{argmin}}_{x\in[a,b]} f(x) \quad \iff \quad f'(u)(v-u) \ge 0 \quad \forall v \in[a,b]$$
\end{itemize}

\vspace{-5mm}
\begin{center}
\only<1>{\vspace{20.5mm}}

\includegraphics<2>[height=25mm]{../talk-oxford/images/calcone.png}
\end{center}
\end{frame}

@@

\section{nonlinear multigrid for PDEs}

\subsection{full approximation scheme (FAS)}

\begin{frame}{nonlinear 2-mesh scheme for PDEs}

@@

	\begin{itemize}
	\item[$\circ$] $\mathcal{V}=H^1(\Omega)$, $F : \mathcal{V} \to \mathcal{V}'$ continuous, $\ell\in \mathcal{V}'$
	\item[$\circ$] example is the Liouville-Bratu problem: $-\grad^2 u - e^u = f$
	\end{itemize}
\item discretization gives algebraic system on fine mesh $\Omega^h$:
    $${\color{FireBrick} F^h(u^h) = \ell^h}$$
\end{itemize}

@@

	\begin{itemize}
	\item are there alternatives to Newton's method?
	\end{itemize}
\item notes:
    \setbeamertemplate{enumerate item}{\emph{\roman{enumi})}}
    \begin{enumerate}
    \item \emph{residual} $r^h = \ell^h - F^h(w^h)$ is computable but \emph{error} $e^h = u^h-w^h$ is unknown
    \item equation can be rewritten as \qquad $F^h(u^h) - F^h(w^h) = r^h$
    \end{enumerate}
\end{itemize}
}
\only<3>{
\begin{itemize}
\item \emph{goal}: from iterate $w^h$, to solve {\color{FireBrick} $F^h(u^h) - F^h(w^h) = r^h$}
\item for $F^h$ linear, one may convert to the error equation
    $$F^h(e^h) = r^h$$
with approximate solution $\tilde e^h$, to improve the iterate
    $$w^h \leftarrow w^h+\tilde e^h$$
\end{itemize}

\vspace{5mm}
}
\only<4>{
\begin{itemize}
\item \emph{updated goal}: use a coarser mesh $\Omega^H$ to somehow estimate the solution $u^h$ in the nonlinear \emph{correction equation}
    $${\color{FireBrick} F^h(u^h) - F^h(w^h) = r^h}$$
\item \emph{smoothers}: there are inexpensive algorithms which improve $w^h$ by efficiently removing high-frequency error components
	\begin{itemize}
	\item sweeping through and solving nodewise problems (nonlinear Gauss-Seidel) is a smoother
	\end{itemize}
\item how about low-frequency error components?
\end{itemize}

\vspace{5mm}
}
\only<5>{
\begin{itemize}
\item Brandt's (1977) \emph{full approximation scheme} (FAS) equation:
	$${\color{FireBrick} F^H(u^H) - F^H(\iR w^h) = R \, r^h(w^h)}$$

@@

\item rewritten FAS equation: let $\ell^H = F^H(\iR w^h) + R\, r^h(w^h)$ then
    $${\color{FireBrick} F^H(u^H) = \ell^H}$$
\end{itemize}
}
\end{frame}

\begin{frame}{full approximation scheme (FAS): 2-mesh cycle}

@@

\begin{align*}
&\text{pre-smooth over fine:} & & \text{smoother on } w^h \\
&\text{restrict:}                   & &\ell^H = F^H(\iR w^h) + R\, r^h(w^h) \\
&\text{solve coarse:}                      & &F^H(w^H) = \ell^H \\
&\text{correct:}                    & &w^h \leftarrow w^h + P(w^H - \iR w^h) \\
&\text{post-smooth over fine:} & & \text{smoother on } w^h
\end{align*}

\bigskip
{\small
\begin{itemize}
\item $R: (\mathcal{V}^h)' \to (\mathcal{V}^H)'$ is \emph{canonical restriction}
\item $P: \mathcal{V}^H \to \mathcal{V}^h$ is \emph{canonical prolongation}
\item $\iR: \mathcal{V}^h \to \mathcal{V}^H$ is \emph{injection}
\item \textbf{restrict}$+$(\textbf{solve coarse})$+$\textbf{correct} \, $=$ \, \emph{FAS coarse grid correction}
\end{itemize}
}
\end{frame}

@@

\begin{frame}{FAS reputation}

\begin{itemize}
\item FAS multigrid works \alert{very well} on nice nonlinear PDE problems
\item example: Liouville-Bratu equation
    $$-\nabla^2 u - e^u = 0$$
with Dirichlet boundary conditions on $\Omega=(0,1)^2$
\item minimal problem-specific code:
    \begin{enumerate}
    \item[1.] residual evaluation on grid level: $F^j(\cdot)$
    \item[2.] pointwise smoother: $\phi_i(c) = 0 \,\forall i$
        \begin{itemize}
        \item[$\circ$] nonlinear Gauss-Seidel iteration
        \end{itemize}
    \item[3.] coarsest-level solve can be same as smoother
    \end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}{the meanings of ``fast solver''}

\begin{itemize}
\item what does ``very well'' on the previous slide mean?
\begin{block}{definition} a solver is \emph{optimal} if work is $O(N)$ for $N$ unknowns
\end{block}
    \begin{itemize}
    \item[$\circ$] measure work in flops or run-time
    \item[$\circ$] constant iterations of $O(N)$ method $\implies$ optimal
    \end{itemize}
\end{itemize}

\begin{itemize}
\item multigrid people get greedy:
\begin{block}{definition} a solver shows \emph{textbook multigrid efficiency} if it does total work less than 10 times that of a single smoother sweep
\end{block}
    \begin{itemize}
    \item[$\circ$] TME $\implies$ optimal
    \end{itemize}
\end{itemize}
\end{frame}

\section{FASCD = FAS multigrid for VIs using constraint decomposition}

\subsection{results}  % FIXME lead with results

\subsection{algorithm details}

@@

% FIXME move earlier \section{results}
